---
title: "Include custom validation functions"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(hubValidations)
```


Custom validation functions can be included and configured within standard `hubValidation` workflows by **including a `validations.yml` file in the `hub-config` directory**. Alternatively, an appropriately structured file can be included at a different location and the path to the file provided through argument `validations_cfg_path`.

`hubValidations` uses the [`config`](https://rstudio.github.io/config/articles/inheritance.html) package to get validation configuration. This allows for configuration inheritance and the ability to include executable R code. 
See the `confog` package vignette on [inheritance and R expressions](https://rstudio.github.io/config/articles/inheritance.html) for more details.

## `validation.yml` structure

`validation.yml` files should follow the nested structure described below:

### Default configuration

The top of any `validation.yml` file, under the required `default:` top level property, should contain default custom validation configurations that will be executed regardless of round ID.

Within the default configuration, individual checks can be configured for each of the 3 validation functions run as part of `validate_submission()`, using the following structure for each validation function:

- **`<name-of-caller-function>`:** One of `validate_model_data`, `validate_model_metadata` and `validate_model_file` depending on the function the custom check is to be included in.
    - **`<name-of-check>`:** The name of the check. This is the name of the element containing the result of the check when `hub_validations` is returned (required).
        - **`fn`:** The name of the check function to be run, as character string (required).
        - **`pkg`:** The name of the package namespace from which to get check function. Must be supplied if function is distributed as part of a package.
        - **`source:`** Path to `.R` script containing function code to be sourced. If relative, should be relative to the hub's directory root. Must be supplied if function is not part of a package and only exists as a script.
        - **`args`:** A yaml dictionary of key/value pairs or arguments to be passed to the custom function. Values can be yaml lists or even executable R code (optional).
        
Note that each of the `validate_*()` functions contain a standard objects in their call environment which are passed automatically to any custom check function and therefore do not need including in the `args` configuration. 

- **`validate_model_file`:**
  - `file_path`: character string of path to file being validated relative to the `model-output` directory.
  - `hub_path`: character string of path to hub.
  - `round_id`: character string of `round_id`
  - `file_meta`: named list containing `round_id`, `team_abbr`, `model_abbr` and `model_id` details.
- **`validate_model_data`:** 
  - `tbl`: a tibble of the model output data being validated.
  - `file_path`: character string of path to file being validated relative to the `model-output` directory.
  - `hub_path`: character string of path to hub.
  - `round_id`: character string of `round_id`
  - `file_meta`: named list containing `round_id`, `team_abbr`, `model_abbr` and `model_id` details.
  - `round_id_col`: character string of name of `tbl` column containing `round_id` information.
- **`validate_model_metadata`:** 
  - `file_path`: character string of path to file being validated relative to the `model-output` directory.
  - `hub_path`: character string of path to hub.
  - `round_id`: character string of `round_id`
  - `file_meta`: named list containing `round_id`, `team_abbr`, `model_abbr` and `model_id` details.

The `args` configuration can be used to override objects from the caller environment.


Here's an example configuration for a single check (`cfg_check_tbl_col_timediff()`) to be run as part of the `validate_model_data()` validation function which checks the content of the model data submission files.

```{r, eval=FALSE, code=readLines(system.file('testhubs/flusight/hub-config/validations.yml', package = 'hubValidations'))}

```

### Round specific configuration

Additional round specific configurations can be included in `validation.yml` that can add to or override default configurations.

For example, in the following `validation.yml`, if the file being validated is being submitted to a round with round ID `"2023-08-15"`, default `col_timediff` check configuration will be overiden by the `2023-08-15` configuration.

```yml
default:
    validate_model_data:
      col_timediff:
        fn: "cfg_check_tbl_col_timediff"
        pkg: "hubValidations"
        args:
          t0_colname: "forecast_date"
          t1_colname: "target_end_date"
          timediff: !expr lubridate::weeks(2)

2023-08-15:
    validate_model_data:
      col_timediff:
        fn: "cfg_check_tbl_col_timediff"
        pkg: "hubValidations"
        args:
          t0_colname: "forecast_date"
          t1_colname: "target_end_date"
          timediff: !expr lubridate::weeks(1)
```

## Managing dependencies of custom sourced functions

TODO
