---
title: "Deploying custom validation functions"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(hubValidations)
```


Custom validation functions can be included and configured within standard `hubValidation` workflows by **including a `validations.yml` file in the `hub-config` directory**. Alternatively, an appropriately structured file can be included at a different location and the path to the file provided through argument `validations_cfg_path`.

`hubValidations` uses the [`config`](https://rstudio.github.io/config/articles/inheritance.html) package to get validation configuration. This allows for configuration inheritance and the ability to include executable R code. 
See the `confog` package vignette on [inheritance and R expressions](https://rstudio.github.io/config/articles/inheritance.html) for more details.

## `validations.yml` structure

`validations.yml` files should follow the nested structure described below:

### Default configuration

The top of any `validations.yml` file, under the required `default:` top level property, should contain default custom validation configurations that will be executed regardless of round ID.

Within the default configuration, individual checks can be configured for each of the 3 validation functions run as part of `validate_submission()`, using the following structure for each validation function:

- **`<name-of-caller-function>`:** One of `validate_model_data`, `validate_model_metadata` and `validate_model_file` depending on the function the custom check is to be included in.
    - **`<name-of-check>`:** The name of the check. This is the name of the element containing the result of the check when `hub_validations` is returned (required).
        - **`fn`:** The name of the check function to be run, as character string (required).
        - **`pkg`:** The name of the package namespace from which to get check function. Must be supplied if function is distributed as part of a package.
        - **`source:`** Path to `.R` script containing function code to be sourced. If relative, should be relative to the hub's directory root. Must be supplied if function is not part of a package and only exists as a script.
        - **`args`:** A yaml dictionary of key/value pairs or arguments to be passed to the custom function. Values can be yaml lists or even executable R code (optional).
        

```{r child="children/_custom-fn-available-args.Rmd", echo=FALSE, results="asis"}
```


Here's an example configuration for a single check (`opt_check_tbl_horizon_timediff()`) to be run as part of the `validate_model_data()` validation function which checks the content of the model data submission files.

```{r, eval=FALSE, code=readLines(system.file('testhubs/flusight/hub-config/validations.yml', package = 'hubValidations'))}
```

The above configuration file relies on default values for arguments `horizon_colname` (`"horizon"`) and `timediff` (`lubridate::weeks()`). We can use the `validations.yml` `args` list to override the default values. Here's an example that includes **executable r code** as the value of an argument.

```
default:
    validate_model_data:
      horizon_timediff:
        fn: "opt_check_tbl_horizon_timediff"
        pkg: "hubValidations"
        args:
          t0_colname: "forecast_date"
          t1_colname: "target_end_date"
          horizon_colname: "horizons"
          timediff: !expr lubridate::weeks(2)
```

### Round specific configuration

Additional round specific configurations can be included in `validations.yml` that can add to or override default configurations.

For example, in the following `validations.yml` which deploys the `opt_check_tbl_col_timediff()` optional check, if the file being validated is being submitted to a round with round ID `"2023-08-15"`, default `col_timediff` check configuration will be overridden by the `2023-08-15` configuration.

```yml
default:
    validate_model_data:
      col_timediff:
        fn: "opt_check_tbl_col_timediff"
        pkg: "hubValidations"
        args:
          t0_colname: "forecast_date"
          t1_colname: "target_end_date"

2023-08-15:
    validate_model_data:
      col_timediff:
        fn: "opt_check_tbl_col_timediff"
        pkg: "hubValidations"
        args:
          t0_colname: "forecast_date"
          t1_colname: "target_end_date"
          timediff: !expr lubridate::weeks(1)
```

## Available optional functions

`hubValidations` includes a number of optional checks or checks that require administrator configuration to be run, detailed below.

For more detail on each function and its configuration parameters, consult the function documentation.

### For deploying through `validate_model_data`

```{r, echo=FALSE}
library(kableExtra)
arrow::read_csv_arrow(system.file("check_table.csv", package = "hubValidations")) %>%
  dplyr::filter(
    .data$`parent fun` == "validate_model_data",
    .data$optional
  ) %>%
  dplyr::select(-"parent fun", -"optional", -"Name") %>%
  dplyr::select("check fun", dplyr::everything()) %>%
  dplyr::mutate("Extra info" = dplyr::case_when(
    is.na(.data$`Extra info`) ~ "",
    TRUE ~ .data$`Extra info`
  )) %>%
  knitr::kable(
    caption =
      "Details of available optional checks or checks requiring configuration for `validate_model_data()`."
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  column_spec(1, bold = TRUE)
```

### For deploying through `validate_model_metadata`

```{r, echo=FALSE}
library(kableExtra)
arrow::read_csv_arrow(system.file("check_table.csv", package = "hubValidations")) %>%
  dplyr::filter(
    .data$`parent fun` == "validate_model_metadata",
    .data$optional
  ) %>%
  dplyr::select(-"parent fun", -"optional", -"Name") %>%
  dplyr::select("check fun", dplyr::everything()) %>%
  dplyr::mutate("Extra info" = dplyr::case_when(
    is.na(.data$`Extra info`) ~ "",
    TRUE ~ .data$`Extra info`
  )) %>%
  knitr::kable(
    caption =
      "Details of available optional checks or checks requiring configuration for `validate_model_metadata()`."
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  column_spec(1, bold = TRUE)
```


## Managing dependencies of custom sourced functions

TODO
