---
title: "Structure of hub_validations class objects"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(hubValidations)
```

The high level `validate_*()` family of functions all return a `<hub_validations>` S3 class object.

## Structure of `<hub_validations>` object

A `hub_validations` object is effectively a list and represents the collected output of the series of checks performed by a higher level `validate_*()` function.

Each named element of the list contains the result of an individual check and inherits from subclass `<hub_check>`. 
The name of each element is the name of the check.

Let's examine an example output of a model output file validation using `validate_submission()`.

```{r}
hub_path <- system.file("testhubs/simple", package = "hubValidations")

v <- validate_submission(hub_path,
  file_path = "team1-goodmodel/2022-10-08-team1-goodmodel.csv"
)

str(v, max.level = 1)
```

The super class returned in each element depends on the status of the check:

- If a check succeeds, a `<message/check_success>` condition class object is returned.

- If a check is skipped, a `<message/check_info>` condition class object is returned.

- Checks vary with respect to whether they return an `<error/check_error>` or `<error/check_failure>` condition class object if the check fails. 
Ultimately, both will cause overall validation to fail and the two classes are used primarily to communicate the severity of a failing check.

## `hub_validations` print method

`hub_validations` objects have their own print method which displays the result, the file name and message of each check:

- `✔` indicates a check was successful (a `<message/check_success>` condition class object was returned)
- `✖` indicates a high severity check failed (a `<error/check_error>` condition class object was returned)
- `!` indicates a lower severity check failed (a `<error/check_failure>` condition class object was returned)
- `ℹ` indicates a check was skipped (a `<message/check_info>` condition class object was returned)

```{r}
v
```

## Structure of a `<hub_check>` object

Let's look more closely at the structure of the first few elements of the `hub_validations` object retuned by `validate_submission()`

```{r}
v <- validate_submission(hub_path,
  file_path = "team1-goodmodel/2022-10-08-team1-goodmodel.csv"
)

str(utils::head(v))
```

Each `<hub_check>` objects contains the following elements:

- `message`: the result message containing details about the check.
- `where:`: there the check was performed, usually the model output file name.
- `call`: the function used to perform the check.
- `use_cli_format`: whether the message is formatted using cli format, almost always TRUE.

## Extra information

Some `<hub_check>` objects contain extra information about the failing check to help identify affected rows in submissions.

For example, the `<hub_check>` object returned for the `valid_vals` check, which checks that all columns in a model output file (excluding the `value` column) contain valid combinations of task ID / output type / output type ID values contains an additional element called `error_tbl`, with details of the invalid value combinations in the rows affected.

To access `error_tbl` from the output of `validate_submission()` stored in an object `v`, you would use:

```{r, eval=FALSE}
v$valid_vals$error_tbl
```

