---
title: "Validating Target Data Pull Requests on GitHub"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(hubValidations)
```

## Running validation checks on a Pull Request with `validate_target_pr()`

The `validate_target_pr()` function is designed to validate target data contributions through Pull Requests on GitHub.

Target data files are individually validated using `validate_target_submission()` on each file, and the affected target type datasets as a whole are validated via `validate_target_dataset()`.
Hub config files are also validated as part of the checks.
Any other files included in the PR are ignored but flagged in a message.

(_See the end of this article for details of the standard checks performed on each file. For more information on deploying optional or custom functions please check the article on [including custom functions](deploying-custom-functions.html) (`vignette("articles/deploying-custom-functions")`)_).

### Deploying `validate_target_pr()` through a GitHub Action workflow

The most common way to deploy `validate_target_pr()` is through a GitHub Action that triggers when a pull request containing changes to target data files is opened.
The hubverse maintains the [**`validate-target-data.yaml`**](https://github.com/hubverse-org/hubverse-actions/tree/main/validate-target-data) GitHub Action workflow template for deploying `validate_target_pr()`.

The latest release of the workflow can be added to a hub's GitHub Action workflows using the `hubCI` package:
```{r, eval = FALSE}
hubCI::use_hub_github_action("validate-target-data")
```

The pertinent section of the workflow is:

```yaml
      - name: Run validations
        env:
          PR_NUMBER: ${{ github.event.number }}
        run: |
          library("hubValidations")
          v <- hubValidations::validate_target_pr(
              gh_repo = Sys.getenv("GITHUB_REPOSITORY"),
              pr_number = Sys.getenv("PR_NUMBER")
          )
          hubValidations::check_for_errors(v, verbose = TRUE)
        shell: Rscript {0}
```

where `validate_target_pr()` is called on the contents of the current Pull Request, the results (an S3 `<target_validations>` class object) are stored in `v` and then `check_for_errors()` is used to signal whether overall validations have passed or failed and summarise any validation failures.

### Example: Validating a Pull Request

Here's an example of validating a PR that adds a valid oracle-output file:

```{r}
tmp_dir <- withr::local_tempdir()
ci_target_hub_path <- fs::path(tmp_dir, "target")
gert::git_clone(
  url = "https://github.com/hubverse-org/ci-testhub-target.git",
  path = ci_target_hub_path
)
gert::git_branch_checkout("add-file-oracle-output", repo = ci_target_hub_path)

v <- validate_target_pr(
  hub_path = ci_target_hub_path,
  gh_repo = "hubverse-org/ci-testhub-target",
  pr_number = 1
)
v
```

```{r}
check_for_errors(v)
```

### Configuring target data validation

For the most robust validation, hubs should include a `target-data.json` configuration file in their `hub-config` directory. When present, this config provides deterministic validation by explicitly defining the date column, column names and types, and observable unit structure.

For hubs without a `target-data.json` config, the **`date_col`** parameter can be used to specify the name of the column containing observation dates (e.g., `"target_end_date"`). This is important for correct schema creation, particularly when the date column is also used for partitioning. When `target-data.json` exists, any user-provided `date_col` is ignored in favour of the config value.

### Relaxed date validation for time-series data

By default, date values in time-series target data are validated strictly against the dates defined in `tasks.json`. However, target data often contains historical observations with dates beyond the hub's configured modeling rounds.

Setting **`allow_extra_dates = TRUE`** relaxes date validation for time-series data, allowing historical observations while still strictly validating other task ID values. Oracle-output data always uses strict date validation regardless of this setting.

```yaml
      - name: Run validations
        env:
          PR_NUMBER: ${{ github.event.number }}
        run: |
          library("hubValidations")
          v <- hubValidations::validate_target_pr(
              gh_repo = Sys.getenv("GITHUB_REPOSITORY"),
              pr_number = Sys.getenv("PR_NUMBER"),
              allow_extra_dates = TRUE
          )
          hubValidations::check_for_errors(v, verbose = TRUE)
        shell: Rscript {0}
```

### Configuring file modification/deletion checks

By default, `validate_target_pr()` does **not** perform modification or deletion checks on target data files.
This reflects the expectation that target data may be updated or corrected over time.

This behaviour can be modified through the `file_modification_check` argument, which controls whether modification/deletion checks are performed and what is returned if modifications/deletions are detected:

- **`"none"`** (default): No modification/deletion checks performed.
- **`"message"`**: Appends a `<message/check_info>` condition class object for each applicable modified/deleted file. Will not result in validation workflow failure.
- **`"failure"`**: Appends a `<error/check_failure>` condition class object for each applicable modified/deleted file. Will result in validation workflow failure.
- **`"error"`**: Appends a `<error/check_error>` condition class object for each applicable modified/deleted file. Will result in validation workflow failure.

Here's an example of a PR that deletes some target data files. With `file_modification_check = "failure"`, this produces a validation failure:

```{r, error=TRUE}
gert::git_branch_checkout("delete-target-dir-files", repo = ci_target_hub_path)

v_mod <- validate_target_pr(
  hub_path = ci_target_hub_path,
  gh_repo = "hubverse-org/ci-testhub-target",
  pr_number = 5,
  file_modification_check = "failure"
)
v_mod

check_for_errors(v_mod)
```

### Allowing deletion of entire target type datasets

By default, deletion of an entire target type dataset (i.e., all files of a target type) in a PR is not allowed.
This can be changed by setting `allow_target_type_deletion` to `TRUE`.

Here's an example of a PR that removes the time-series dataset and adds oracle-output data. With the default settings, this produces an error:

```{r, error=TRUE}
gert::git_branch_checkout("remove-ts-add-oo", repo = ci_target_hub_path)

v_del <- validate_target_pr(
  hub_path = ci_target_hub_path,
  gh_repo = "hubverse-org/ci-testhub-target",
  pr_number = 4
)
v_del

check_for_errors(v_del)
```

Setting `allow_target_type_deletion = TRUE` allows the deletion to pass:

```{r}
v_del_allowed <- validate_target_pr(
  hub_path = ci_target_hub_path,
  gh_repo = "hubverse-org/ci-testhub-target",
  pr_number = 4,
  allow_target_type_deletion = TRUE
)
check_for_errors(v_del_allowed)
```

## Checking for validation failures with `check_for_errors()`

`check_for_errors()` is used to inspect a `target_validations` class object, determine whether overall validations have passed or failed and summarise any detected errors/failures.

### Validation failure

If any elements of the `target_validations` object contain `<error/check_error>`, `<warning/check_warning>` or `<error/check_exec_error>` condition class objects, the function throws an error and prints the messages from the failing checks.

### Validation success

If all validation checks pass, `check_for_errors()` returns `TRUE` silently and prints:

```
All validation checks have been successful.
```

### Verbose output

If printing the results of all checks is preferred instead of just summarising the results of checks that failed, argument `verbose` can be set to `TRUE`.


## `validate_target_pr` check details

For details on the structure of `<hub_validations>` objects, including on how to access more information about specific checks, see `vignette("articles/hub-validations-class")`.

### Checks on target datasets

```{r, echo=FALSE}
library(kableExtra)
arrow::read_csv_arrow(system.file("check_table.csv", package = "hubValidations"))  |>
  dplyr::filter(
    .data$`parent fun` == "validate_target_dataset" |
      .data$`check fun` == "check_config_hub_valid",
    !.data$optional
  )  |>
  dplyr::select(-"parent fun", -"check fun", -"optional")  |>
  dplyr::mutate("Extra info" = dplyr::case_when(
    is.na(.data$`Extra info`) ~ "",
    TRUE ~ .data$`Extra info`
  ))  |>
  knitr::kable(caption = "Details of dataset-level checks performed by `validate_target_pr()`.")  |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))  |>
  column_spec(1, bold = TRUE)
```


### Checks on individual target files

```{r, echo=FALSE}
arrow::read_csv_arrow(system.file("check_table.csv", package = "hubValidations"))  |>
  dplyr::filter(
    .data$`parent fun` %in% c(
      "validate_target_file",
      "validate_target_data"
    ),
    !.data$optional
  )  |>
  dplyr::select(-"parent fun", -"check fun", -"optional")  |>
  dplyr::mutate("Extra info" = dplyr::case_when(
    is.na(.data$`Extra info`) ~ "",
    TRUE ~ .data$`Extra info`
  ))  |>

  knitr::kable(
    caption = "Details of file-level checks performed by `validate_target_submission()`."
  )  |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))  |>
  column_spec(1, bold = TRUE)
```


<div class="alert alert-info" role="note">

#### Custom checks

The standard checks discussed here are the checks deployed by default by the `validate_target_pr` function. For more information on deploying optional or custom functions please check the article on [deploying custom functions](deploying-custom-functions.html) (`vignette("articles/deploying-custom-functions")`).

</div>
