---
title: "Writing custom validation functions"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

r_path <- function(hub_path, file_name) {
  fs::path(hub_path, "src", "validations", "R", file_name, ext = "R")
}
```

```{r setup}
library(hubValidations)
```

While `hubValidations` provides a wide range of validation `check_*()` functions, there are times when you might need to write your own custom check functions to check a specific aspect of your hub's submissions.  

This guide will help you understand how to write custom check functions and what tools are available in `hubValidations` to help. 

While more details about deploying custom check functions during validation workflows are available in **`vignette("deploying-custom-functions")`**, it's useful to mention here that custom functions are configured through the `validations.yml` file and executed as part of `validate_model_data()`, `validate_model_metadata()` and `validate_model_file()` functions.

# Anatomy of a check function

While source code of existing `hubValidations` `check_*()` functions can be a good place to start when writing custom check functions, it is important to understand the structure of a check function, particularly the expected inputs and outputs.

At it's most basic, a custom check function should:

- evaluate whether a condition is met
- return an appropriate check condition object

In addition, if the check condition is not met, it's also helpful to capture any details that can guide users towards specifics of the failure and how to fix it

In general, `hubValidations` check functions evaluate conditions with respect to one or more of the following:

- Model output submission files
- Model output submission file content (i.e data)
- Model metadata files


## `create_custom_check()` for creating custom check function templates

To help you get started on the right path, we also provide function `create_custom_check()` for creating a basic custom check function from a template. 

The function requires a name for the new custom check function, e.g. `"example_check"`. It then creates an `.R` script file named after the function (`example_check.R`) and saves it in the hub at the recommended location: `src/validations/R/`. The script contains basic skeleton code to create a custom check function called `example_check`.

The output of `create_custom_check()` can also be parametarised through a number of arguments to include additional template code snippets (see below for examples).

Let's take a look at the basic structure of a custom check function created by `create_custom_check()`:


```{r}
hub_path <- withr::local_tempdir()

create_custom_check("cstm_check_tbl_basic",
  hub_path = hub_path
)
```

The contents of the created file at `src/validations/R/cstm_check_tbl_basic.R` are as follows:


```{r, code=readLines(r_path(hub_path, "cstm_check_tbl_basic")), eval=FALSE}
```


<div class="alert alert-info">

**Function naming conventions**

While not necessary for deploying custom functions, we recommend following the naming conventions used in `hubValidations`.

- `check_file_*` for checks that operate on model output files (e.g. file location, name etc).
- `check_tbl_*` for checks that operate on a tibble of model output data (i.e. the contents of a file).
- `check_meta_*` for checks that operate on model metadata files.
- `check_submission_*` for checks that operate on high level properties of the submission (e.g. timing).
- `check_config_*` for checks that operate on the hub configuration files.
- `check_valid_*` for checks that don't fit into the above categories.

To distinguish `hubValidations` package functions from custom ones, we also recommend prefixing custom function names with an additional prefix, e.g. `cstm_` or `cs`.

</div>

## Arguments to custom check functions

### Arguments available in the caller environment 

The `hubValidations` package automatically passes a set of standard objects to custom check functions. These objects are passed to arguments of custom check functions with the same name and do not need to be included in the `args` configuration in the `validations.yml` file.


```{r child="children/_custom-fn-available-args.Rmd", echo=FALSE, results="asis"}
```

### Additional arguments

You can add additional arguments to custom check functions and pass values to them by including them in the `args` configuration in the `validations.yml` file. These values are passed to the custom check function by `hubValidations` when the function is called.


If you do add additional arguments to a custom check function, you should also add input  checks at the start of the function to ensure inputs are valid. [The `checkmate` package](https://mllg.github.io/checkmate/) contains a wide range of functions for checking inputs.

For example, the optional check `opt_check_tbl_col_timediff()` (which is deployed in exactly the same fashion as custom functions, i.e. through the `validations.yml` file) takes additional arguments `t0_colname`, `t1_colname` and `timediff`. 


```{r}
opt_check_tbl_col_timediff
```



You can add an example extra argument with `extra_args = TRUE` when creating the custom check function with `create_custom_check()`.

```{r}
create_custom_check("cstm_check_tbl_args",
  hub_path = hub_path,
  extra_args = TRUE
)
```

This adds an extra example argument `extra_arg` to the custom check function formals as well as an example input check to the top of the function body.

```{r, code=readLines(r_path(hub_path, "cstm_check_tbl_args")), eval=FALSE}
```


## Function output

### Capturing and returning check results

**Returning a check condition with `capture_check_cnd()`**

The `capture_check_cnd()` function is used to return a check condition and it's output should also be what a custom check function returns in most cases. The function returns a `<hub_check>` class object depending on the value passed to the `check` argument, which represents the summary of the condition being checked by a given validation function. 

If the value passed to `check ` is `TRUE`, the function returns a `<message/check_success>` class object. 

If the value is `FALSE`, the output depends on the `error` argument.  

- If `error` is `TRUE`, the function returns a `<message/check_error>` class object. If `error` is `FALSE`, the function returns a `<error/check_failure>` class object, which indicates the check has failed. 
- If `error` is `TRUE`, the function returns a `<error/check_error>` class object, which indicates the check has failed and additionally causes execution of further custom validation functions to halt. Set `error = TRUE` if downstream checks cannot be run if the current check fails.

The default value of `error` is `FALSE` in `create_custom_check()` but setting it to `TRUE` sets the error argument to `TRUE`.

```{r}
create_custom_check("cstm_check_tbl_error",
  hub_path = hub_path, error = TRUE
)
```

```{r, code=readLines(r_path(hub_path, "cstm_check_tbl_error")), eval=FALSE}
```


### Skipping a check

**Returning a check message with `capture_check_info()`**

Sometimes a check function might not always be applicable and a pre-condition needs to be met before the main check itself is performed. **If the pre-condition is not met, the check is usually skipped**. 

**For such checks, the function should return a `<message/check_info>` object**, generated by the `capture_check_info()` function. Use the `msg` argument to explain that a check was skipped and why.


```{r}
capture_check_info(
  "modelA-teamA/2024-09-12-modelA-teamA",
  "Condition for running this check was not met. Skipped."
)
```

For example, the `check_tbl_value_col_ascending()` check function which validates that values are ascending when arranged by increasing `output_type_id` order is only applicable to `cdf` and `quantile` output types. Before proceeding with the main check, the function first checks whether the model output `tbl` contains data for `cdf` and `quantile` output types. If not, the check is skipped.

```{r}
check_tbl_value_col_ascending
```


You can add a pre-condition check block of code with argument `conditional = TRUE` when creating the custom check function with `create_custom_check()`.

```{r}
create_custom_check("cstm_check_tbl_skip",
  hub_path = hub_path,
  conditional = TRUE
)
```


```{r, code=readLines(r_path(hub_path, "cstm_check_tbl_skip")), eval=FALSE}
```


## Loading config files

Many check conditions against information stored in hub configuration files and these are often read in at the start of the custom check function.

The easiest way to make hub configuration information available within your function is to pass the `hub_path` caller environment object by specifying it as a function argument and then use `hubUtils::read_config(hub_path)` to read in the `tasks.json` configuration file.

You can add a `config = TRUE` argument when creating the custom check function with `create_custom_check()` to include the `hub_path` argument and insert a code snippet in you custom check function skeleton that reads in the `tasks.json` hub configuration file.


```{r}
create_custom_check("cstm_check_tbl_config",
  hub_path = hub_path,
  config = TRUE
)
```


```{r, code=readLines(r_path(hub_path, "cstm_check_tbl_config")), eval=FALSE}
```


# Custom function dependencies

When writing your functions you might want to use functions from other packages. 


```{r child="children/_add-deps-source.Rmd", echo=FALSE, results="asis"}
```

 
# Managing custom check functions as a package

The simplest way to manage custom check functions is to store them as scripts in the `src/validations/R` directory in the root of the hub and source them during validation by specifying the path to a custom functions file via the `source:` property in `validations.yml`.

Alternatively, you could **manage your custom functions as a package**. 

You can easily [turn the contents of `src/validations` into a local `validations` R package](https://r-pkgs.org/whole-game.html#create_package) with:

```{r, eval=FALSE}
usethis::create_package("src/validations", open = FALSE)
```

This would allow you to:

- Make your functions available locally by users who could use `pak::local_install("src/validations")` in the hub root to install the `validations` package.
- [Manage additional dependencies](https://r-pkgs.org/description.html#sec-description-imports-suggests) required by your custom functions formally through the `DESCRIPTION` file.
- [Formally test your custom functions](https://r-pkgs.org/testing-basics.html) using `testthat` tests in the `tests/testthat` directory.


```{r child="children/_add-deps-pkg.Rmd", echo=FALSE, results="asis"}
```

If you want to share custom functions across multiple hubs, you could even consider separating them into a standalone package and hosting them on GitHub.
